Crystal as a thermodynamic computer

### 2024-09-12 paper/main.tex
updat

With the proposed Ising model, we are interested in how efficiently we can drive the system to the ground state, which encodes the output of the CA.
In contrast with a traditional computation model, cooling an energy model does not have a specific direction.
It can be done from the input to the output, from the output to the input or neither.
Intuitively, cooling the deterministic computation is much easier than cooling the non-deterministic computation.
Since the latter encodes the circuit satisfaction problem, which is NP-hard, while the former is a simple linear-time computation.
However, this is not always the case.
Global cooling by gradually lowering the temperature (or annealing), as in the Ising machine~\cite{fabre2014optical}, may still hard for a deterministic computation, no matter how large the input detuning $\Delta_{in}$ is.

@@

\subsubsection{Simulated annealing algorithm}
Since the natural cooling process is highly nontrivial to model, we consider a theoretical model for cooling, which is the simulated annealing (SA)~\cite{Kirkpatrick1983}.
When the energy model in \Cref{eq:ising} thermalizes at temperature $T$, the spin configuration $\vs = (s_1, s_2, \cdots, s_n)$ follows the Boltzmann distribution:
\begin{equation}
    p(\vs) = \frac{1}{Z}\exp\left(- \frac{E({\vs})}{k_bT}\right),
\end{equation}
where $Z = \sum_{\vs} \exp\left(-\frac{E({\vs})}{k_bT}\right)$ is the partition function, 
$E({\vs})$ is the energy of the spin configuration $\vs$, and $k_b$ is the Boltzmann constant, which is set $k_b = 1$ in this work.
%We employ the Metropolis-Hastings algorithm~\cite{metropolis1953equation} to characterize the thermodynamic process of the TC system as modulated by the heat bath.
SA is a Monte Carlo method~\cite{metropolis1953equation} that simulates the thermal annealing process.
It performs a random walk in the state space of the TC, each step we try to flip a spin according to some unbiased prior distribution,
and the probability of accepting a move from state $\vs$ to $\vs^\prime$ is given by:
\begin{equation}
    P_{\vs^\prime\leftarrow \vs} = \frac{1}{1+\exp\left(\frac{E({\vs^\prime}) - E({\vs})}{T}\right)}.
\end{equation}
This accepting probability follows from the Glauber dynamics~\cite{glauber1963time}.
It can be shown that this process is Markovian and ergodic, and the system will eventually reach the thermal equilibrium.
The speed of the convergence is determined by the spectral gap of the transition matrix of the Markov chain, which is defined as
\begin{equation}
    \Delta = 1 - \frac{\lambda_2}{\lambda_1},
\end{equation}
where $\lambda_1$ and $\lambda_2$ are the largest and second-largest eigenvalues of the transition matrix, respectively.
In order to sample from the distribution $\epsilon$ close to the equilibrium distribution, the number of steps required is $O(\log_{1-\Delta} \epsilon)$, i.e. the distribution quality improves exponentially with the number of sampling steps.

\subsection{Cooling computation with an idealized heat bath}\label{cooling-computation-with-moving-heat-bath}

@@

The computational process of the TC is equivalent to finding the ground state of the TC with the spin configuration in the left surface being fixed.

We focus here on the thermodynamics-driven approach to finding ground states, 
specifically the simulated annealing (SA) method. Traditional SA method can be seen as the whole system interacting with a heat bath, by gradually decreasing the temperature
of the heat bath the TC can be driven to the ground state if the temperature is decreased slowly enough.
However, the cooling process can last for exponential time for some hard spin-glass problems.

We introduce a heat zone of width $W$, which represents the region of the system that interacts with the heat bath, as shown in \Cref{Surface-programmable-thermodynamic computer}.
Let layer $i$ be defined as the set of spins with x-coordinate $i$.
The system is devided into $\lceil\frac{M}{W}\rceil$ parts,
where part $j$ consists of layers in the range $[jW, (j+1)W-1]$.
The system can then be cooled by sequentially cooling each parts with an idealized heat bath, 
which maintains a space-invariant temperature within the heat zone while keeping other regions at zero temperature.
To drive the system toward the ground state, the cooling order should proceed from left to right
, corresponding to the computation direction of the cellular automata.

@@

Now we turn to determining the sweep time $t_{th}$ required to achieve an error probability $\epsilon$ in each part. 
Let $\lambda_1$ and $\lambda_2$ be the largest and second largest eigenvalues of the Markov transition matrix, respectively. 
The relationship between $\frac{\lambda_2}{\lambda_1}$ and $W$ can be expressed as
\begin{equation}
    \label{eq:spectral-gap}
    1 - \frac{\lambda_2}{\lambda_1} \sim e^{-W}
\end{equation}

Consequently, a relationship exists between $t_{th}$ and $\epsilon$, which is given by:

\begin{equation}
    \left( \frac{\lambda_2}{\lambda_1} \right) ^ {t_{th}} = \epsilon \rightarrow t_{th} \sim \ln(\frac{M}{W})e^{W}
\end{equation}

And the total sweep time scale as $\frac{M}{W}\ln(\frac{M}{W})e^W$.

@@

\subsection{Cooling computation with a realistic heat bath}\label{estimation-of-the-computing-time}

@@

\subsubsection{Testing model with non-equilibrium simulated
annealing}\label{some-details-about-non-equilibrium-simulated-annealing}

@@

There are also Monte Carlo-based methods to simulate a moving
temperature wave packet that facilitates material growth~\cite{Godfrey1995, Tan2017}.
However, these methods fail to establish a direct connection between the simulated temperature and experimental
temperature~\cite{Zollner2014}.
Therefore, based on HeatBath acceptance rule, we made a slight modification to the classical SA model.

@@

>>>>

With the proposed Ising model, we are interested in how efficiently we can drive the system to the ground state, which encodes the output of the CA.
In contrast with a traditional computation model, cooling an energy model does not have a specific direction.
It can be done from the input to the output, from the output to the input or neither.
Intuitively, cooling the deterministic computation is much easier than cooling the non-deterministic computation.
Since the latter encodes the circuit satisfaction problem, which is NP-hard, while the former is a simple linear-time computation.
However, this is not always the case.
Global cooling by gradually lowering the temperature (or annealing), as in the Ising machine~\cite{fabre2014optical}, may still hard for a deterministic computation, no matter how large the input detuning $\Delta_{in}$ is.
It can be rigorous proved that the cooling process is hard if the energy landscape of the optimization problem exhibits the overlap gap property~\cite{Gamarnik2021}. If a problem has the overlap gap property, then it can not be solved by any local algorithms in time polynomial to the system size, including both the physical annealing and the simulated annealing method used in this work.

@@

Since there is no standard approach to model the physical annealing process, we consider a theoretical model for cooling, which is the simulated annealing (SA)~\cite{Kirkpatrick1983} algorithm that detailed in \Cref{app:sa}.

\subsection{Dynamic annealing}\label{cooling-computation-with-moving-heat-bath}
\jinguo{Please shorten this section!}

@@

To overcome the overlap gap property, we propose to cool the system piece by piece, by introducing a heat zone with width $W$.
It represents the region of the system that interacts with the heat bath, as shown in \Cref{Surface-programmable-thermodynamic computer}.
Let layer $i$ be defined as the set of spins with x-coordinate $i$.
The system is devided into $\lceil\frac{M}{W}\rceil$ parts,
where part $j$ consists of layers in the range $[jW, (j+1)W-1]$.
The system can then be cooled by sequentially cooling each parts with an idealized heat bath, 
which maintains a space-invariant temperature within the heat zone while keeping other regions at zero temperature.
To drive the system toward the ground state, the cooling order should proceed from left to right
, corresponding to the computation direction of the cellular automata.

@@

Now we turn to determining the sweep time $t_{th}$ required to achieve an error probability $\epsilon$ in each part. 
Let $\lambda_1$ and $\lambda_2$ be the largest and second largest eigenvalues of the Markov transition matrix, respectively. 
We assume the relationship between $\frac{\lambda_2}{\lambda_1}$ and $W$ can be expressed as
\begin{equation}
    \label{eq:spectral-gap}
    1 - \frac{\lambda_2}{\lambda_1} \sim e^{-W}.
\end{equation}
Consequently, $t_{th}$ is related to $\epsilon$ by:
\begin{equation}
    \left( \frac{\lambda_2}{\lambda_1} \right) ^ {t_{th}} = \epsilon \rightarrow t_{th} \sim \ln(\frac{M}{W})e^{W},
\end{equation}
and the total sweep times scale as $\frac{M}{W}\ln(\frac{M}{W})e^W$.

@@

\subsubsection{A more realistic heat bath}\label{estimation-of-the-computing-time}

@@

\subsubsection{Testing model with non-equilibrium simulated annealing}\label{some-details-about-non-equilibrium-simulated-annealing}\jinguo{Suggested subtitle: Simulating temperature gradient}

@@

There are also Monte Carlo-based methods to simulate a moving
temperature wave packet that facilitates material growth~\cite{Godfrey1995, Tan2017}.
However, these methods fail to establish a direct connection between the simulated temperature and experimental
temperature~\cite{Zollner2014}.
Therefore, based on HeatBath acceptance rule, we made a slight modification to the classical SA model.\jinguo{??? shouldn't it be Glauber dynamics?}

@@

\section{Simulated annealing algorithm}\label{app:sa}
When the energy model in \Cref{eq:ising} thermalizes at temperature $T$, the spin configuration $\vs = (s_1, s_2, \cdots, s_n)$ follows the Boltzmann distribution:
\begin{equation}
    p(\vs) = \frac{1}{Z}\exp\left(- \frac{E({\vs})}{k_bT}\right),
\end{equation}
where $Z = \sum_{\vs} \exp\left(-\frac{E({\vs})}{k_bT}\right)$ is the partition function, 
$E({\vs})$ is the energy of the spin configuration $\vs$, and $k_b$ is the Boltzmann constant, which is set $k_b = 1$ in this work.

%We employ the Metropolis-Hastings algorithm~\cite{metropolis1953equation} to characterize the thermodynamic process of the TC system as modulated by the heat bath.
SA is a Monte Carlo method~\cite{metropolis1953equation} that simulates the thermal annealing process.
It performs a random walk in the state space of the TC, each step we try to flip a spin according to some unbiased prior distribution,
and the probability of accepting a move from state $\vs$ to $\vs^\prime$ is given by:
\begin{equation}
    P_{\vs^\prime\leftarrow \vs} = \frac{1}{1+\exp\left(\frac{E({\vs^\prime}) - E({\vs})}{T}\right)}.
\end{equation}
This accepting probability follows from the Glauber dynamics~\cite{glauber1963time}.
It can be shown that this process is Markovian and ergodic, and the system will eventually reach the thermal equilibrium.
The speed of the convergence is determined by the spectral gap of the transition matrix of the Markov chain, which is defined as
\begin{equation}
    \Delta = 1 - \frac{\lambda_2}{\lambda_1},
\end{equation}
where $\lambda_1$ and $\lambda_2$ are the largest and second-largest eigenvalues of the transition matrix, respectively.
In order to sample from the distribution $\epsilon$ close to the equilibrium distribution, the number of steps required is $O(\log_{1-\Delta} \epsilon)$, i.e. the distribution quality improves exponentially with the number of sampling steps.
